Trade-off between market and ecosystem services drives settlement decisions among smallholder ranchers in Baja California Sur, Mexico

### 2024-02-19 manuscript/manuscript.qmd
additional round of reviews and revisions

In BCS (along with the rest of Mexico), the Agrarian Land Reforms of 1992 played
an outsized role in driving these changes, as they introduced powerful new
incentives for *ejidos* to parcelize collective land holds and allocate private
rights to individuals [@barnes2009]. While not the sole cause, this new property
regime has increased uncertainty and insecurity about land title and exposed the
region to speculation from real estate markets. As has happened elsewhere, these
privatization efforts appear to be led by NGOs, commercial interests, and other
external forces, with the result being greater fragmentation of pastoral
rangeland systems and increased uncertainty about land use
[@burnsilver_etal2008; @galvin2009]. For the rural ranchers of the Sierra de la
Giganta, this process of fragmentation and increasing insecurity at least
partially explains why ranches of related individuals tend to cluster.
Rancheros, unable to find new land for themselves, simply build ranches next to
their kin on land that is at least historically associated with their family.

To summarize, we want to know how traditional rural ranchers in BCS navigate the
competing costs and benefits of market integration and local ecosystem services.
This is done by estimating their contribution within a regression framework to
the location and size of ranch clusters in two east-west running watersheds
along the southern extent of the Sierra de la Giganta range. Given the extreme
ecological constraints that characterize this region, especially the aridity of
the region, we expect water management to play a substantial role in individual
settlement decisions. Privatization may impose an additional constraint on this
system, especially as it has served to increase land insecurity.

@@

To evaluate the determinants of ranch occupation and population size, we rely on
two generalized linear models (GLMs). We use the first GLM to evaluate whether
market integration or local ecology has played a larger role in structuring the
distribution of occupied and abandoned ranch clusters. For this model we assume
that the dependent variable, whether a ranch cluster is occupied or abandoned,
has a Bernoulli distribution and is related to the linear predictor by the logit
link: $$
\begin{aligned}
Y &\sim B(1, p) \\
E[Y] &= p \\
logit(p) &= \beta_{0} + \beta_{1} D(La\; Paz) + \beta_{2} D(Springs)
\end{aligned}
$$ with $p$ being the probability that a ranch cluster is occupied and $D(x)$
the travel time from a ranch cluster to $x$ (see below for more details about
how we calculate this).

We use the second GLM to identify which variable contributes more to the
distribution of rancher populations across currently occupied clusters, with an
additional variable representing land security, or the proportion of ranch
houses within a cluster having clear title to their property, as this may
mediate ranch population size. For this model, we assume that the dependent
variable, the population size of a ranch cluster, has a negative binomial
distribution and is related to the linear predictor by the log link: $$
\begin{aligned}
Y &\sim NB(\lambda, k) \\
E[Y] &= \lambda \\
log(\lambda) &= \beta_{0} + \beta_{1} D(Cities) + \beta_{2}D(Springs) + \beta_{3} P(Security)
\end{aligned}
$$ with $\lambda$ being the average number of individuals per ranch cluster, $k$
a dispersion parameter, and $P(Security)$ the proportion of ranches in a ranch
cluster that are secure. The use of a negative binomial rather than the standard
Poisson model for count data is preferred here to account for dispersion, or
circumstances in which the variance deviates from the mean, which violates a key
assumption of the Poisson distribution.

@@

The independent variables, market integration and habitat suitability, we
operationalize using three estimates of travel time: (i) one to La Paz, (ii) one
to Ciudad Constitucion or La Paz (depending on the watershed), and (iii) one to
local springs, with each involving a mix of pedestrian and automotive travel.
The first two variables are our proxies for market integration, under the
assumption that increased interaction with markets will positively covary with
proximity to the cities. The third variable, travel time to springs, is our
proxy for habitat suitability, as water is a critical resource in this arid
environment. In the binomial GLM, which examines the distribution of abandoned
versus active ranching clusters, we use only travel time to La Paz (i) as Ciudad
Constitucion did not offer a significant market when those locations were being
established. We then use travel time to both Ciudad Constitucion and La Paz (ii)
in the negative binomial GLM examining ranch-cluster population sizes. travel
time to springs (iii) is used in both models, on the assumption that the
location of active streams would not have changed during the period of time in
question [@grismer1993].

@@

We initiated this research to understand how local ecology and market
integration influenced settlement dynamics. Our first model examines how these
two factors influence whether or not a ranch cluster is occupied or abandoned.
The model suggests that a cluster's status (whether it is occupied or abandoned)
negatively co-varies with travel time to spring ($\beta$ = -4.803, *p* = 0.009)
and positively co-varies with travel time to La Paz ($\beta$ = 2.797, *p* \<
0.001), indicating that ranch clusters are more likely to be occupied closer to
springs and further from town (@tbl-coefficients, @fig-response-plots). Our
second model examines how local ecology and market integration influence the
distribution of the population across occupied ranch clusters today. The model
shows that population size across occupied ranch clusters negatively co-varies
with travel time to cities ($\beta$ = -0.802, *p* \< 0.001), for both
Constitucion and La Paz, but does not significantly co-vary with travel time to
springs ($\beta$ = 0.699, *p* = 0.299) (@tbl-coefficients, @fig-response-plots).
The model with an additive land security term suggests that more secure ranch
clusters have slightly smaller populations, but the effect is non-significant
($\beta$ = -0.114, *p* = 0.687). The ANOVA also indicates that an additive land
security term does not significantly improve the model ($\chi$(1) = 3.003, *p* =
0.684). The same is true for an interactive term ($\chi$(1) = 3.130, *p* =
0.663).

@@

One important advantage of this research concerns the use of travel time as a
measure of both market integration and local adaptation. Using this simple
metric has a number of advantages. First, it provides a straightforward way of
evaluating the relative contribution of each variable to the outcome of
interest, in this case, settlement decisions among rural ranchers. Second, it
conforms with geographic theory regarding spatial interactions and dependence
(otherwise known as Tobler's Law, @tobler1970) and overcomes known limitations
of Euclidean distance as a cost measure. Finally, it is a relatively simple
matter to derive those cost estimates from easily accessible remote-sensed
elevation and road network data held in online repositories. High resolution
informant data, on the other hand, are generally difficult and costly to
acquire, often with little chance of providing the right sort of overlap between
different variables required for fair comparisons.

It is true, as @mattison2022 point out, that market integration is a
multi-dimensional process, making it at first blush an awkward concept to try
and capture in a single variable like travel time or caloric-intake
[@henrich2010]. Still, Tobler's Law would seem to suggest that all the variables
strongly associated with market integration will also co-vary with proximity to
cities and their markets. In other words, the costs and benefits of market
integration will positively co-vary with access to markets - both will increase
as one gains greater access to and interacts more often with markets. So, the
closer individuals are, the more likely they should be to derive their caloric
requirements from cities, to work in cities, to own televisions, cellphones, and
other goods sold in cities, but also the more likely they should be to face
intense competition for jobs, higher rents, and other costs associated with
living in that proximity.

@@

Our measure of land security proved to be non-significant in our model. Rather
than being an outcome from which to draw conclusions, we believe this result
should be interpreted with caution. Our measure of land security is based on a
combination of government data and informant observations and so likely does not
capture the full complexity of this important issue. A larger dataset with a
more standardized and quantifiable measure of land security would be more
representative. We also note that the issue of land security is not equally
salient to everyone in this population. Many of the informants that might be
land insecure by most objective measures do not necessarily act as if their
title to land is uncertain. This would suggest that perceptions of risk with
respect to property rights should be studied more thoroughly.

@@

The findings presented here have important consequences for traditional ranching
communities living in similar circumstances, for how they maintain their way of
life while meeting the challenges posed by climate change, land fragmentation,
and other exogenous forces. For instance, researchers have shown that market
integration within traditional agricultural communities involves a shift from
intensive to extensive kinship systems, or shifts from small kin networks with
higher degrees of relatedness to large kin networks with lower degrees of
relatedness [@peña2002]. This appears to be the result of a decreasing reliance
on consanguineous or close-kin marriage customs, with families favoring
expansive social networks within a market economy over stronger family ties
within a largely agrarian economy [@shenk2016]. Currently, the ranching
communities of BCS are characterized by more intensive kin networks, but the
limited available evidence suggests that this might be changing, with
individuals increasingly marrying outside of the ranching community, most likely
to individuals living within the cities. Families are as a consequence
increasing their connections to the cities and by extension increasing the
incentives for integrating more closely with their markets.

>>>>

In BCS (along with the rest of Mexico), the Agrarian Land Reforms of 1992 played
an outsized role in driving these changes, as they introduced powerful new
incentives for *ejidos* to parcelize collective land holdings and allocate
private rights to individuals [@barnes2009]. While not the sole cause, this new
property regime has increased uncertainty and insecurity about land title and
exposed the region to speculation from real estate markets. As has happened
elsewhere, these privatization efforts appear to be led by NGOs, commercial
interests, and other external forces, with the result being greater
fragmentation of pastoral rangeland systems and increased uncertainty about land
use [@burnsilver_etal2008; @galvin2009]. For the rural ranchers of the Sierra de
la Giganta, this process of fragmentation and increasing insecurity at least
partially explains why ranches of related individuals tend to cluster.
Rancheros, unable to find new land for themselves, simply build ranches next to
their kin on land that is at least historically associated with their family.

To summarize, we want to know how traditional rural ranchers in BCS navigate the
competing costs and benefits of market integration and local ecosystem services.
This is done by estimating their contribution within a regression framework to
the location and size of ranch clusters in two east-west running watersheds
along the southern extent of the Sierra de la Giganta range. Given the extreme
ecological constraints that characterize this region, especially the aridity of
the region, we expect water resources to play a substantial role in individual
settlement decisions. Privatization may impose an additional constraint on this
system, especially as it has served to increase land insecurity.

@@

To evaluate the determinants of ranch occupation and population size, we rely on
two generalized linear models (GLMs). We use the first GLM to evaluate whether
market integration or local ecology has played a larger role in structuring the
distribution of occupied and abandoned ranch clusters. For this model we assume
that the dependent variable, whether a ranch cluster is occupied or abandoned,
has a Bernoulli distribution and is related to the linear predictor by the logit
link:$$
\begin{aligned}
Y &\sim B(1, p) \\
E[Y] &= p \\
logit(p) &= \beta_{0} + \beta_{1} D(La\; Paz) + \beta_{2} D(Springs)
\end{aligned}
$$ {#eq-logit-model} with $p$ being the probability that a ranch cluster is
occupied and $D(x)$ the travel time from a ranch cluster to $x$ (see below for
more details about how we calculate this).

We use the second GLM to identify which variable contributes more to the
distribution of rancher populations across currently occupied clusters, with an
additional variable representing land security, or the proportion of ranch
houses within a cluster having clear title to their property, as this may
mediate ranch population size. For this model, we assume that the dependent
variable, the population size of a ranch cluster, has a negative binomial
distribution and is related to the linear predictor by the log link:$$
\begin{aligned}
Y &\sim NB(\lambda, k) \\
E[Y] &= \lambda \\
log(\lambda) &= \beta_{0} + \beta_{1} D(Cities) + \beta_{2}D(Springs) + \beta_{3} P(Security)
\end{aligned}
$$ {#eq-poisson-model} with $\lambda$ being the average number of individuals
per ranch cluster, $k$ a dispersion parameter, and $P(Security)$ the proportion
of ranches in a ranch cluster that are secure. The use of a negative binomial
rather than the standard Poisson model for count data is preferred here to
account for dispersion, or circumstances in which the variance deviates from the
mean, which violates a key assumption of the Poisson distribution.

@@

The independent variables, market integration and habitat suitability, we
operationalize using three estimates of travel time: (i) one to La Paz, (ii) one
to Ciudad Constitucion or La Paz (depending on the watershed), and (iii) one to
local springs, with each involving a mix of pedestrian and automotive travel.
The first two variables are our proxies for market integration, under the
assumption that increased interaction with markets will positively covary with
proximity to the cities. The third variable, travel time to springs, is our
proxy for habitat suitability, as water is a critical resource in this arid
environment. In the binomial GLM, which examines the distribution of abandoned
versus active ranching clusters, we use only travel time to La Paz (i) as Ciudad
Constitucion did not offer a significant market when those locations were being
established. We then use travel time to both Ciudad Constitucion and La Paz (ii)
in the negative binomial GLM examining ranch-cluster population sizes. Travel
time to springs (iii) is used in both models, on the assumption that the
location of active streams would not have changed during the period of time in
question [@grismer1993].

@@

We initiated this research to understand how local ecology and market
integration influenced settlement dynamics. Our first model examines how these
two factors influence whether or not a ranch cluster is occupied or abandoned.
The model suggests that a cluster's status (whether it is occupied or abandoned)
negatively co-varies with travel time to spring ($\beta$ = -4.803, *p* = 0.009)
and positively co-varies with travel time to La Paz ($\beta$ = 2.797, *p* \<
0.001), indicating that ranch clusters are more likely to be occupied closer to
springs and further from town (@tbl-coefficients, @fig-response-plots). Our
second model examines how local ecology and market integration influence the
distribution of the population across occupied ranch clusters today. The model
shows that population size across occupied ranch clusters negatively co-varies
with travel time to cities ($\beta$ = -0.802, *p* \< 0.001), for both
Constitucion and La Paz, but is not significantly correlated with travel time to
springs ($\beta$ = 0.699, *p* = 0.299) (@tbl-coefficients, @fig-response-plots).
The model with an additive land security term suggests that more secure ranch
clusters have slightly smaller populations, but the effect is non-significant
($\beta$ = -0.114, *p* = 0.687). The ANOVA also indicates that an additive land
security term does not significantly improve the model ($\chi$(1) = 3.003, *p* =
0.684). The same is true for an interactive term ($\chi$(1) = 3.130, *p* =
0.663).

@@

One important advantage of this research concerns the use of travel time as a
measure of both market integration and local adaptation. Using this simple
metric has a number of advantages. First, it is corroborated by previous
research [@johansson_etal2002; @corsi_etal2017]. Second, it provides a
straightforward way of evaluating the relative contribution of each variable to
the outcome of interest, in this case, settlement decisions among rural
ranchers. Third, it conforms with geographic theory regarding spatial
interactions and dependence (otherwise known as Tobler's Law, [@tobler1970]) and
overcomes known limitations of Euclidean distance as a cost measure. Finally, it
is a relatively simple matter to derive those cost estimates from easily
accessible remote-sensed elevation and road network data held in online
repositories. High resolution informant data, on the other hand, are generally
difficult and costly to acquire, often with little chance of providing the right
sort of overlap between different variables required for fair comparisons.

It is true, as @mattison2022 point out, that market integration is a
multi-dimensional process, making it at first blush an awkward concept to try
and capture in a single variable like travel time or caloric-intake
[@henrich2010]. Still, Tobler's Law would seem to suggest that all the variables
strongly associated with market integration will also be correlated with
proximity to cities and their markets. In other words, the costs and benefits of
market integration will be positively correlated with access to markets - both
will increase as one gains greater access to and interacts more often with
markets. So, the closer individuals are, the more likely they should be to
derive their caloric requirements from cities, to work in cities, to own
televisions, cellphones, and other goods sold in cities, but also the more
likely they should be to face intense competition for jobs, higher rents, and
other costs associated with living in that proximity.

@@

Our measure of land security proved to be non-significant in our model. Rather
than being an outcome from which to draw conclusions, we believe this result
should be interpreted with caution as previous scholarship suggests that land
tenure is essential for securing the livelihood of smallholders
[@corsi_etal2017; @besley_ghatak2010; @wickeri_kalhan2010;
@aggarwal_freudenberger2013; @dejanvry_etal2015] and ensuring sustainable
resource management. Furthermore, our research on the effect of land security on
ranching decision-making in this population suggests that it influences whether
or not households engage in cattle production -- a risky economic decision in
this arid ecosystem [@macfarlan_etal2023a]. So, what explains our negative
result? Our argument is threefold. First, the distribution of land secure and
insecure households is a result of changes to the property regime that happened
in the 1990s [@dejanvry_etal2015], around the time when the distribution of
ranches and ranching families was established, so there may be a lag effect that
our analysis is insensitive to. With additional data on the distribution of
households across a larger spatial area (or temporal window, for the
establishment of new households into the future), we may find that land
insecurity is having a real effect on individual choices. Second, our measure of
land security is based on a combination of government data and informant
observations and so likely does not capture the full complexity of this
important issue. A larger dataset with a more standardized and quantifiable
measure of land security would be more representative. Finally, we note that the
issue of land security, particularly as that is understood as clear title, is
not equally salient to everyone in this population [@lambrecht_asare2016]. Many
of the informants that might be land insecure by most objective measures do not
necessarily act as if their title to land is uncertain. This would suggest that
perceptions of risk with respect to property rights should be studied more
thoroughly.

@@

The findings presented here have important consequences for traditional ranching
communities living in similar circumstances, for how they maintain their way of
life while meeting the challenges posed by climate change, land fragmentation,
and other exogenous forces. For instance, researchers have shown that market
integration within traditional agricultural communities involves a shift from
intensive to extensive kinship systems, or shifts from small kin networks with
higher degrees of relatedness to large kin networks with lower degrees of
relatedness [@peña2002]. This appears to be the result of a decreasing reliance
on consanguineous or close-kin marriage customs, with families favoring
expansive social networks within a market economy over stronger family ties
within a largely agrarian economy [@shenk2016]. Currently, the ranching
communities of BCS are characterized by more intensive kin networks
[@macfarlan_etal2020b], but the limited available evidence suggests that this
might be changing, with individuals increasingly marrying outside of the
ranching community, most likely to individuals living within the cities
[@page_etal2024]. Families are as a consequence increasing their connections to
the cities and by extension increasing the incentives for integrating more
closely with their markets.
